# 2.0  Regression and model validation

*Describe the work you have done this week and summarize your learning.*

# Describe your work and results clearly. 
# Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
# Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
setwd("~/IODS-project")
library(dplyr)
learning2014 <- readxl::read_excel("~/IODS-project/learnings2014.xlsx")
str(learning2014) #structure of data
dim(learning2014) #Dimension of data
names("learning2014")
library(dplyr)
````

```{r}
learning2014 <- readxl::read_excel("~/IODS-project/data/learning2014.xlsx") %>%
  mutate_at(vars(gender), factor)
#c("gender")
pairs(learning2014[-1], col=learning2014$gender)
#pairs(learning2014[!names(learning2014) %in% c("gender")],col=learning2014$gender)
#install.packages("GGally",repos = c("http://rstudio.org/_packages","http://cran.rstudio.com")

#install.packages("ggplot2",repos = c("http://rstudio.org/_packages","http://cran.rstudio.com")
# access the GGally and ggplot2 libraries

````

# create a more advanced plot matrix with ggpairs()
```{r}
library(GGally)
library(ggplot2)
````



```{r}
#a scatter plot of point versus attitude


qplot(attitude, points, data = learning2014) + geom_smooth(method = "lm")
# fit a linear model
my_model <- lm(points ~ attitude, data = learning2014)

# print out a summary of the model
summary(my_model)

# an plot matrix with ggpairs()
ggpairs(learning2014, lower = list(combo = wrap("facethist", bins = 20)))

#a regression model with multiple explanatory variables
my_model2 <- lm(points ~ attitude + stra + surf, data = learning2014)

#summary of the model
summary(my_model2)

# draw diagnostic plots using the plot() function. Choose the plots 1, 2 and 5
par(mfrow = c(2,2))
plot(my_model2, which = c(1,2,5))
```` 

