---
editor_options: Jussi Pekka Vehviläinen
markdown: `date()`
wrap: 72
---

# IODS 2022 Exercise set 2
This week topic was Regression and model validation. Here we present basic R commands to build up summary tables and linear models which can be used for finding cause-and-effect relationships between different variables.

# Data wrangling

## PART 1

```{r}
# Set up workind dir
setwd("D://IODS-project") 
dir.create(paste0(getwd(),"/Data"))
setwd(paste0(getwd(),"/Data"))
```

## PART 2
Bring .txt file to R environment which has been downloaded to Data - folder.
```{r}
learning_2014<-as.data.frame(read.table("Data/learning2014.txt", header = TRUE, sep = "\t"))
```

Table should have 183 rows and 60 cols containing integer values in all except last column which contains information about gender by character.

## PART 3

-   Subset Data set to contain columns: gender, age, attitude, deep, stra, surf and points.
-   Remove also rows which have 0 in Points.
-   Deep is calculated by taking the mean of cols: c("D03","D11","D19","D27","D03","D11","D19","D27","D06","D15","D23","D31") and excluding 0.
-   Stra is calculated by taking the mean of cols: c("ST01","ST09","ST17","ST25","ST04","ST12","ST20","ST28") and excluding 0.
-   Surf is calculated by taking the mean of cols: c("SU02","SU10","SU18","SU26","SU05","SU13","SU21","SU29","SU08","SU16","SU24","SU32") and excluding 0.

```{r}
# Remove rows which have 0 in Points
analy_dataset<-subset(learning_2014, as.numeric(Points)!=0)
analy_dataset[analy_dataset==0]<-NA

# Build new dataframe (DF) for needed data. Gender, Age and Points are added straigth but for start Deep, Stra and Surf has been set to NA.
df<-data.frame(Gender=analy_dataset$gender,Age=analy_dataset$Age,Attitude=NA, Deep=NA,Stra=NA, Surf=NA, Points=analy_dataset$Points)

# Add Deep to DF by taking the mean

deep <- c("D03","D11","D19","D27","D03","D11","D19","D27","D06","D15","D23","D31")
df$Deep<-rowMeans(analy_dataset[,colnames(analy_dataset) %in% deep], na.rm=TRUE)

# Add Stra to DF by taking the mean
stra <- c("ST01","ST09","ST17","ST25","ST04","ST12","ST20","ST28")
df$Stra<-rowMeans(analy_dataset[,colnames(analy_dataset) %in% stra], na.rm=TRUE)

# Add Surf to DF by taking the mean
surf <- c("SU02","SU10","SU18","SU26","SU05","SU13","SU21","SU29","SU08","SU16","SU24","SU32")
df$Surf<-rowMeans(analy_dataset[,colnames(analy_dataset) %in% surf], na.rm=TRUE)

# Add Attitude to DF by taking the mean
attitude <- c("Da","Db","Dc","Dd","De","Df","Dg","Dh","Di","Dj")
df$Attitude<-rowMeans(analy_dataset[,colnames(analy_dataset) %in% attitude], na.rm=TRUE)
```

## PART 4

```{r}
# Change working dir
setwd("D://IODS-project")
# Write results to .csv format file
write.csv2(df,"Data/learning2014.csv")
# Read saved .csv file. Header= TRUE means that file has headers. sep=";" means that values are separated by ;. dec="," means that in the file ',' has been used for decimals
learning_2014.2<-as.data.frame(read.table("Data/learning2014.csv", header = TRUE, sep = ";", dec=","))
```

# Analysis

## PART 1 - Read file to R environment

Data set contains summary results of course 'Johdatus yhteiskuntatilastotieteeseen, syksy 2014' survey. There should be 7 variables (gender, age, attitude, deep, stra, surf and points) and 166 observations.''' Gender: Male = 1 Female = 2 Age: Age (in years) derived from the date of birth Attitude: Global attitude toward statistics. Mean of original variables (\~Da+Db+Dc+Dd+De+Df+Dg+Dh+Di+Dj) Deep: Deep approach. Mean of original variables (\~d_sm+d_ri+d_ue) Stra: Strategic approach. Mean of original variables ( \~st_os+st_tm) Surf: Surface approach. Mean of original variables (\~su_lp+su_um+su_sb) Points: Total counts from survey. More information about used variables can be found from <http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS2-meta.txt>'''

```{r}
# Read file to R environment
learning_2014.2<-as.data.frame(read.table("Data/learning2014.csv", header = TRUE, sep = ";", dec=","))
# Check dimensions of read table
dim(learning_2014.2)
```

## PART 2

By commend summary we can take a look 'inside' the data. By the command we get 'boxplot information' about our numerical data in dataframe. Scatterplot matrix is used to describe relationships between the variables. It's constructed from the dataframe with ggpairs -function (ggplot2 -package). Result plot shows, in addition of variables relationships, variables diverging and gives correlation coefficients with asterix showing level of significance.

```{r}
# Summary
library(ggplot2)
library(GGally)
summary(learning_2014.2[,2:ncol(learning_2014.2)])
# Scatterplot
p <- ggpairs(learning_2014.2[-(1:2)], mapping = aes(), lower =list(combo = wrap("facethist", bins = 20))) 
print(p)
```

Most promising relationship seems to be between: Attitude and Points, and Surf and Deep.\
There seems to be also some kind of relationship between: Stra and Surf. As overall, matrix gives good overlook of data, and starting point to study more relationships between variables'''

## PART 3 and PART 4

Create a regression model with multiple explanatory variables

```{r}
my_model2 <- lm(Points ~ Attitude + Stra + Surf, data = learning_2014.2)
summary(my_model2)
```

Summary of a regression model shows that only Attitude seems to correlate significantly with Points. From the print, you can see model residiuals; summary table (estimate value, std. error, t-value and p-value for all variables in model against Points). Significance of variable correlation can be read from p-value(last column of Coefficients table). Significance levels threshols are given under the table. Summary gives also p-value for whole model which isn't significant because model contains variables that hasn't have relationship to Points. In next step, lest remove other variables from the model and see what happens

```{r}
# Only Attitude seems to be significant, so lets do model again with only adding it
my_model3 <- lm(Points ~ Attitude, data = learning_2014.2)
summary(my_model3)
```

Now results seems to be better and p-value is significant for the variable relationship as well as for the model. Multiple R-squared is the proportion of the variation in dependent variable that can be explained by the independent variable. So in the model where we haves three variables 20.74 % of the variation in Points can be explained by variables. But now interesting is that Attitude by itself explains 19.06 % of the variation. Showing us that Stra and Surf effect to the points is pretty minimal.

# PART 5

## Study more our model with diagnostic plots

```{r}
plot(my_model3, which=c(1,2,5))
```

From the 'Residual vs leverage' - plot we can check which and how many of observation are influential. In our case data seems good and there isnt any point outside Cook distance lines.

Also 'Residual vs. Fitted' - plot seems good. Data is divided evenly in x - and y-axle.

QQ-plot also indicates goodness of our model. If the points runs out too early from the line, there might be some other variables effecting our relationship more than the Attitude variable. In this case QQplot seems to be really nice, but noT perfect. \`\`\`
