# Week 3: Logistic Regression

## 3.0 Loading packages

```{r}
library("boot")
library("readr")
```

## 3.1 Data wrangling

Data wrangling completed using the UCI Machine Learning Repository (https://archive.ics.uci.edu/ml/datasets.html). csv files downloaded from the repository and final data set "alc.csv" has been exported into the project folder. Rcodes in the learning diary has been updated. 

## 3.1 Creating new R Markdown file

New Rmd file has been created with tittle "chapter3" and now saved in the project folder.

## 3.2 Importing the data set and exploration

```{r}
library(tidyverse)
stu_alc2014 <- read_csv("alc.csv" , show_col_types= FALSE)
spec(stu_alc2014)

#looking at the data
dim(stu_alc2014)
colnames(stu_alc2014)
glimpse(stu_alc2014)
```
## 3.3 Choosing the variables 

```{r}
stu_alc2014_2 <- select(stu_alc2014, absences, G3, age, freetime, high_use)
str(stu_alc2014_2)
colnames(stu_alc2014_2)
```
In above data set we have 370 observations and 35 variables. These data set belong to a survey from questionnaire in two secondary school from Portugal in which different variables, demographics and socio-economic features measures students association with alcohol consumption. Here I choose 4 interesting variable that I think has greater influence in the level of alcohol consumption among these school kids. 
My rationale behind choosing these variables is that certain age groups have higher access to alcohol, ages above 16 lets say can access alcohol easily than ages below 16 so I wish to see the relation here. Also free time activities and amount of free time influences alcohol consumption. Likewise final grade and absences can directly correlate with higher use. So I wish to test my model with these variables.

## 3.4 Exploring and plotting the choosen variable

```{r}
#Let's explore the choose variables using box plots

##Let's see for absences
g1 <- ggplot(stu_alc2014, aes(x = high_use, y = absences))
g1 + geom_boxplot() + ylab("Absences")

##Let's see for G3(Final Grade)
g1 <- ggplot(stu_alc2014, aes(x = high_use, y = G3))
g1 + geom_boxplot() + ylab("Final Grade")

##Let's see for age
g1 <- ggplot(stu_alc2014, aes(x = high_use, y = age))
g1 + geom_boxplot() + ylab("Age")

##And for freetime
g1 <- ggplot(stu_alc2014, aes(x = high_use, y = freetime))
g1 + geom_boxplot() + ylab("Freetime")

```
General overview from the plot infers some association between high alcohol use and absence and also age (holds true). It would be interesting to fit this kind of model to see the relationship. Final grade and alcohol consumption shows some association but there appears to be some difference in their mean for true and false.Free time doesn't seem to have much effect on alcohol consumption.


## 3.4 Logistic regression 

```{r}
##Lets call this model-1 (m1) which explores 4 variable
m1 <- glm(high_use ~ absences + G3 + age + freetime, data = stu_alc2014_2, family = "binomial")
summary(m1)
```
This seems to be an interesting outcome. Fitted model seem to match my above hypothesis based on box plot and distribution for some variable, absences for instant is the most significant and free time is second most significant among other variable. Absence has the highest significance (p = 0.000249) and free time is significant with p = 0.001511. Final grade and age however doesn't have the same lever of significance as other two. Final grade has p = 0.05 which can be considered significant result but comparatively, two other variable stands out the most. 

## 3.5 Power of the model 

```{r}
coef(m1)
OR <- coef(m1) %>% exp
CI <- confint(m1) %>% exp

#Print OR and CI 
cbind(OR, CI)
```
Coefficient and Confidence Interval: Looking at the coefficient, final grade has negative value suggesting an opposite association between higher alcohol consumption which makes sense because higher grade can result in lower alcohol consumption. absences, age and free time shows positive association with free time being the most cause of higher alcohol consumption followed by age and absences. This is also supported by the odds ratio and confidence interval for each tested variable.  
The above hypothesis therefore holds true for most variable. It is quite obviously important to explore other variable to see effect of an additional variable on an outcome through multivariate analysis. 



