## Dimensionality reduction techniques


large selection of statistical methods making the world less messy - we reduce the dimensions to getter a better picture. 




- 

***
<br>
#### The Data wrangling 

<br>
The packages justncase
```{r libraries, results='hide',message=FALSE}
library(psych)
library(pastecs)
library(tidyr)
library(dplyr)
library(GGally)
library(ggpubr)
library(ggplot2)
library(skimr)
library(DataExplorer)
library(stringr)
library(corrplot)
```
<br>
<br>
__A "biref" summary of the steps taken with R scripts displayed here on Data wrangling__


__Reading data__
```{r}
human <- read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human1.txt", sep  =",", header = T)
dim(human)
```
<br>
__Mutate the data (#1)__

```{r}

str(human$GNI)
str_replace(human$gni, pattern=",", replace ="") %>% as.numeric

```


Expolring datasets with R

- head() *displays the first 6 rows by default*
- dim() *displays the type and a preview of all columns as a row*
- glimpse() *display a vertical preview of the datase (pkg dplyr)*
- summary() *displays data type and attributes which are especially useful for numeric attributes*
- skim() *good addition to the summary function (pkg skimr)*
- DataExplorer *fast track exploration try out using line "DataExplorer::create_report() works better in R cause issues with pandoc??* 


<br>
__Exclude the unneeded variables (#2)__
```{r}
keep <- c("Country", "Edu2.FM", "Labo.FM", "Life.Exp", "Edu.Exp", "GNI", "Mat.Mor", "Ado.Birth", "Parli.F")
human <- dplyr::select(human, one_of(keep))
dim(human)
```
<br>
__Remove all rows with missing values (#3)__
```{r}
complete.cases(human)
data.frame(human[-1], comp = complete.cases(human))
human_ <- filter(human, complete.cases(human))
dim(human_)
```

<br>
__Remove observations not relating to countries (#4)__
```{r}
tail(human_, 10)
last <- nrow(human_) - 7
human_ <- human[1:last, ]
dim(human_)
```
<br>
__Define the row names of the data by the country names (#5)__
```{r}
rownames(human_) <- human_$Country
human_ <- human_[-1]
dim(human_)
colnames(human_)
```


__Saving file with rownames__
```{r}
write.csv(human_, file = "humandata.csv", row.names = TRUE)

```

<br>

#### The Data Analysis

<br>
__graphical overview of the data and summaries of the variables__

```{r}
human <- read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human2.txt", header = TRUE, sep = ",")
head(human)
summary(human)
ggpairs(human)
cor(human) %>% corrplot
```

__COMMENTS: The data displays quite obvious negative correlations among maternl morbitity mratio and the educational factors to life expectansy at birth. This data displays similar associations as the data fro WHO reprts very well seen if one uses GapMinder.__ 





<br>
__PCA Analysis__
<br>
__*from non-standardized data*__
```{r}
pca_human1 <- prcomp(human)
s1 <- summary(pca_human1)
#rounding to prencentages
pca_pr1 <- round(100*s1$importance[2,], digits = 1) 
pca_pr1
pc_lab1 <- paste0(names(pca_pr1), " (", pca_pr1, "%)")
biplot(pca_human1, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = pc_lab1[1], ylab = pc_lab1[2])

```


COMMENTS: 

All features are seemingly pointing towards the PC2 -component and thus contributing mainly to it with the raw data.



<br>
__*from standardized data*__
```{r}
human_std <- scale(human)
pca_human2 <- prcomp(human_std)
s2 <- summary(pca_human2)
#roubding to precentages
pca_pr2 <- round(100*s2$importance[2,], digits = 1) 
pca_pr2
pc_lab2 <- paste0(names(pca_pr2), " (", pca_pr2, "%)")
biplot(pca_human2, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = pc_lab2[1], ylab = pc_lab2[2])

```


__COMMENTS: The precentage of females in parliament and labor force factors show closer clustering compared to other factors.__ 




<br>


__MCA Analysis (#5)__
<br>
__Loading data (tea data / FactoMineR) and visualizations and summaries__



This ted dataset is from a package __FactoMiner__.

- [R Documentation](https://www.rdocumentation.org/packages/FactoMineR/versions/1.42)

- [FactoMiner site](http://factominer.free.fr/)

- [Cran site](https://cran.r-project.org/web/packages/FactoMineR/index.html)

- 36 variables from using 18 questions from 300 individuals on tea habbits.


Script: loading data and visualizations.
```{r}
library(FactoMineR)

data(tea)
dim(tea)
head(tea)
#ggpairs(tea)
#cor(tea) %>% corrplot

```





__MCA Analysis of selected variables__

```{r}
# Looking at the selected data
keep_columns <- c("Tea", "How", "lunch", "relaxing", "healthy", "effect.on.health" )
tea_time <- select(tea, one_of(keep_columns))
summary(tea_time)
gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
head(tea_time)
#ggpairs(tea)

```






```{r}
# multiple correspondence analysis
mca <- MCA(tea_time, graph = FALSE)
# summary of the model
summary(mca)
# visualize MCA
plot(mca, invisible=c("ind"), habillage = "quali")

```

COMMENTS: 








