---
title: "Chapter3(2)"
author: "Omolara Mofikoya"
date: "21 11 2019"
output: html_document

---


# 3. Logistic regression
##  3.1 Overview of Data 
```{r include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)


```{r}
alc <- read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/alc.txt",sep = ",",header = TRUE)
names(alc)
````

this data shows student achievement in secondary education of two Portuguese schools in mat and por
```{r}

glimpse(alc)
````
This study is to investigate the relationship between bte high and low alcohol users and how it affect other variables like sex, activities and grade

###   3.2 Graphical Representation of various variable and how it affects high intake of alcohol


```{r}
alc %>% group_by(sex, high_use) %>% summarise(count = n(), mean_grade = mean(G3))
alc %>% group_by(failures, high_use) %>% summarise(count = n(), mean_absences = mean(absences))
alc %>% group_by( high_use, activities) %>% summarise(count = n())
alc %>% group_by(sex, high_use) %>% summarise(count = n(), health = mean(health))
````

From my observation, there are less high alcohol consumers in both male and female , and for the females the high alcohol consumers tend to have an higer grade while it is the opposite for males

Lets initiate aplot for for high users and each of the variable taken

```{r}
g1 <- ggplot(alc, aes(x = high_use, y = G3, col = sex))
g1 + geom_boxplot() + ylab("grade")
g2 <- ggplot(alc, aes(x = high_use, y = absences, col = sex))
g2 + geom_boxplot() +ggtitle("Student absences by alcohol consumption and sex")
g3 <- ggplot(alc, aes(x = high_use, y = activities, col = sex))
g3 + geom_boxplot() +ggtitle("Student activities by alcohol consumption and sex")
g4 <- ggplot(alc, aes(x = high_use, y = health, col = sex))
g4 + geom_boxplot() +ggtitle("Student health by alcohol consumption and sex")

````
From the graph, we can see that high users have more absences days

```{r}
m <- glm(high_use ~ absences + sex + activities + health + G3, data = alc, family = "binomial")
summary(m)
````
###      3.3Logistic Regression
Based on the result from the coefficent, the grades and health is not a good predictor variable because their p  values ig greater than 0.05
for sex for male = -1.03092 + 1.00755 * 1
           female  = -1.03092 + 1.00755 * 0
 The odd ratio of a male having high alcohol intake is over the odds that a female will have an high intake
 For activities, the odd ratio for high user partaking in activities is lower than high users not taking part in activities.
```{r} 
 OR <- coef(m) %>% exp
 CI <- confint(m)%>% exp
cbind(OR, CI)
m <- glm(high_use ~  absences + sex + activities , data = alc, family = "binomial")
probabilities <- predict(m, type = "response")
alc <- mutate(alc, probability = probabilities)
alc <- mutate(alc, prediction = probability > 0.5)
select(alc, absences, sex,activities, high_use, probability, prediction) %>% tail(10)
table(high_use = alc$high_use, prediction = alc$prediction)
g <- ggplot(alc, aes(x = probability, y = high_use, col = prediction))
g + geom_point()
table(high_use = alc$high_use, prediction = alc$prediction)%>% prop.table()%>% addmargins()
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}
loss_func(class = alc$high_use, prob = alc$probability)
````
the error is 24%

