---
title: "Chapter 4"
author: "Omolara Mofikoya"
date: "21 11 2019"
output: html_document
---

# 4. Clustering and Classification
```{r include=FALSE}
library(MASS)
data("Boston")
str(Boston)
dim(Boston)
````

This data set contains 506 rows(observation) and 14 columns(variables). It contains various information about the city of Boston.

## 4.1 An overview of the data
````{r}
pairs(Boston[1:7])
pairs(Boston[8:14])
summary(Boston)
````


````{r}
library(corrplot)
cor_matrix<- cor(Boston)
corrplot(cor_matrix)
````

````{r}
boston_scaled <- scale(Boston)
summary(boston_scaled)

````
Scaling means subtrating from the mean and dividing by the standdard deviation, after scaling, we can see that all the mean variable is 0

Change the data back to data.frame

````{r}

boston_scaled<- as.data.frame(boston_scaled)

summary(boston_scaled$crim)
crime <- quantile(boston_scaled$crim)
crime
crimes <- cut(boston_scaled$crim, breaks = crime, include.lowest = TRUE, labels = c("low", "med_low",  "med_high", "high"))
table(crimes)
boston_scaled <- dplyr::select(boston_scaled, -crim)
boston_scaled <- data.frame(boston_scaled, crimes)
summary(boston_scaled)
````

````{r}
n <- nrow(boston_scaled)
ind <- sample(n, size = n*0.8)
train <- boston_scaled[ind,]
test <- boston_scaled[-ind,]
correct_classes <- test$crimes
test <- dplyr::select(test,-crimes)
dim(test)
````

## 4.2 Linear Discriminate Analysis
```{r}
lda.fit<- lda(crimes ~., data = train)
lda.fit
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "orange", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}
  classes <- as.numeric(train$crime)
  plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)
````
From the result given in the proportion of trace, the seperation achieved by the first discriminant function is 95.03% and the second is 3.89%.
There is a very large overlap between the low, med_low and med_high, high is well seperated fom the others and has some med_high figures overlapping with it.

## 4.3 Predicting the values
```{r}
lda.pred <- predict(lda.fit, newdata = test )
lda.pred$class
tab<- table(correct = correct_classes, predicted = lda.pred$class)
tab
sum(diag(tab))/sum(tab)
````
Our predictor pricted 80 observation correctly out of 104 obersation. The classifify pridiction of the crime rate is 76.47%

## 4.4 K-means clusterring
```{r}
library(MASS)
data("Boston")
boston_scaledd <- scale(Boston)
boston_scaledd <- as.data.frame(boston_scaledd)
dist_eu<- dist(boston_scaledd)
summary(dist_eu)
dist_man <- dist(Boston, method = 'manhattan')
summary(dist_man)
km <- kmeans(Boston, centers = 3)
pairs(Boston[1:7], col = km$cluster)
pairs(Boston[8:14], col = km$cluster)
````

